The Argennon Virtual Machine

### 2021-04-21 tex/avm.tex
Redesign chapter one
Signed-off-by: aybehrouz <behrouz_ayati@yahoo.com>

%! Author = aybehrouz
%! Date = 2/28/21

% Preamble
\documentclass[a4paper]{report}

@@

    \chapter{Specification}\label{ch:specification}

@@

    The Argennon Virtual Machine (AVM) is an abstract computing machine for executing Argennon's smart contracts. It
    is designed in a way that it could be efficiently implemented in either hardware or software.

    \section{The Structure of the Argennon Virtual Machine}\label{sec:the-structure-of-the-algorand-virtual-machine}

    \subsection{Data Types}\label{subsec:data-types}

@@

    The instruction set of the Argennon Virtual Machine distinguishes its operand types using instructions intended to
    operate on values of specific types. For instance, \texttt{iadd} assumes that its operands are two 64-bit integers.

    \subsection{The \texttt{PC} Register}\label{subsec:the-pc-register}

    \subsection{Call Stack}\label{subsec:call-stack}

    A call stack contains the information that is needed for restoring the state of the invoker of a method.

    \subsection{Memory Unit}\label{subsec:memory-unit}

    The Argennon Virtual Machine has a byte addressable memory which is divided into separate segments. Every segment
    belongs to a smart contract that has a unique \texttt{applicationID}. The AVM always has a single working memory
    segment and memory locations outside its current working segment can not be accessed. The only instructions which
    can change the current working segment are \texttt{invoke\_external}, \texttt{athrow} and return instructions.

    Every smart contract has its own memory segment. Hence, there is no way for a smart contract to access another
    smart contract's memory. Interaction between smart contracts is done using \texttt{invoke\_external} instruction,
    and a smart contract can invoke methods of another smart contract by this instruction.

    A memory segment consists of five data areas:
    \begin{itemize}
        \item Code Area
        \item Constant Area
        \item Local Frame
        \item Operand Stack
        \item Heap
    \end{itemize}
    All data areas except operand stack, have their own address space which starts from 0. Operand stack is a
    last-in-first-out (LIFO) stack and is not addressable. Every memory access instruction operates on its specific
    data areas.

    \subsubsection{Code Area}

    The code area of a segment contains the byte-code of the smart contract which owns that segment. The AVM has no
    instructions for manipulating the code area. \textbf{Installing, removing and updating smart contracts need to
    be done externally.}

@@

    The constant area of a segment contains several kinds of constants, ranging from user defined constants to method
    address tables. A method address table stores method locations in the code area and their access type. The access
    type of a method can be either \texttt{public} or \texttt{private}. The AVM has no instructions for modifying the
    constant area.

    \note{Only public methods can be invoked by \texttt{invoke\_external} instruction.}

@@

    Every time a local frame is created, a corresponding empty last-in-first-out (LIFO) stack is created too. AVM
    instructions take operands from the operand stack, operate on them, and push the result back onto the operand
    stack. An operand stack is destroyed when its owner method completes, whether that completion is normal or abrupt.

@@

    The heap of a segment is a persistent memory area which is divided into pages. Memory locations inside every page
    have a separate address space that starts from 0, and each page can be referenced by an index that starts from 0. In
    other words, the address of every memory location inside a heap is a pair of indices: \texttt{(pageIndex, offset)}.
    Different pages of a heap do not need to be equally sized.

    \note{The reason behind this paged design is that the heap is usually persisted using a block device. A heap with
    a paged structure could expose the underlying block based nature of the persistence layer to the application
    layer. In this way, the compiler or the programmer could better optimize the code for the persistence layer.}

    \section{Instruction Set Summary}\label{sec:instruction-set-summary}

    An Argennon Virtual Machine instruction consists of a \textbf{one-byte} opcode specifying the operation to be
    performed, followed by zero or more operands supplying arguments or data that are used by the operation.
    \textbf{The number and size of the operands are determined solely by the opcode.}

@@

    The Argennon Virtual Machine has three types of method invocation:
    \begin{itemize}
        \item \texttt{invoke\_internal} invokes a method from the current running smart contract.
        \item \texttt{invoke\_external} invokes a \texttt{public} method from another smart contract. It will change
        the current memory segment to the segment of the invoked smart contract.
        \item \texttt{invoke\_native} invokes a method that is not hosted by the Argennon virtual machine. By this
        instruction, high performance native methods of the hosting machine could become available to AVM smart
        contracts.
    \end{itemize}

    \note{In the future, we may need to add special instructions for invoking interface and virtual methods...}

    Each time a method is invoked a new local frame and operand stack is created. The Argennon Virtual Machine uses
    local frames to pass parameters on method invocation. On method invocation, any parameters are passed in
    consecutive local variables stored in the method's local frame starting from local variable 0. The invoker of a
    method writes the parameters in the local frame of the invoked method using \texttt{arg} instructions.

    \subsubsection{Exceptions}

    An exception is thrown programmatically using the \texttt{athrow} instruction. Exceptions can also be thrown by
    various Argennon Virtual Machine instructions if they detect an abnormal condition. Some exceptions are not
    catchable and will always abort the execution of the smart contract.

    \subsubsection{Method Invocation Completion}

    A method invocation completes normally if that invocation does not cause an exception to be thrown, either
    directly from the AVM or as a result of executing an explicit throw statement. If the invocation of the current
    method completes normally, then a value may be returned to the invoking method. This occurs when the invoked
    method executes one of the return instructions, the choice of which must be appropriate for the type of the value
    being returned (if any). Execution then continues normally in the invoking method's local frame with the returned
    value (if any) pushed onto the operand stack.

    A method invocation completes abruptly if an exceptions is thrown and is not caught by the current method. A
    method invocation that completes abruptly never returns a value to its invoker.

    When a method completes, whether normally or abruptly, the call stack is used to restore the state of the invoker,
    including its local frame and operand stack, with the \texttt{PC} register appropriately restored and incremented
    to skip past the method invocation instruction. If the invoker was another smart contract, i.e.\ the invocation
    was made by an \texttt{invoke\_external} instruction, the current memory segment will be changed to the invoker's
    segment.

    A thrown exception causes methods in the call stack to complete \textbf{abruptly} one by one, as long as the
    \texttt{PC} register is not pointing to a \texttt{catch} instruction. The \texttt{catch} instruction acts like a
    branch instruction that branches only if an exception is caught. \textbf{When an external method invocation
    completes abruptly, before changing the current segment, all changes made to the heap area by that method
    call, including changes made to other segments, will be rolled back.} So, when a method call completes abruptly,
    that method call essentially has no effect on the AVM state.

@@

    \subsection{Authorizing Operations}\label{subsec:authorizing-operations}

    In blockchain applications, we usually need to authorize certain operations. For example, for sending an asset
    from a user to another user, first we need to make sure that the sender has authorized this operation. The
    Argennon virtual machine has no built in mechanism for authorizing operations, but it provides a rich set of
    cryptographic instructions for validating signatures and cryptographic entities. By using these instructions and
    passing signatures as parameters to methods, a programmer can implement the required logic for authorizing any
    operation.

@@

    In addition to signatures, a method can verify its invoker by using \texttt{get\_parent} instruction. This
    instruction gets the \texttt{applicationID} of the smart contract that is one level deeper than the current
    smart contract in the call stack. In other words, it returns the \texttt{applicationID} of the smart contract that
    has invoked the current smart contract. (if any)

    \subsection{Heap Allocation Instructions}\label{subsec:heap-allocation-instructions}

@@

    \chapter{Implementation}\label{ch:implementation}

>>>>

%! suppress = TooLargeSection
%! Author = aybehrouz
%! Date = 2/28/21

% Preamble
\documentclass[11pt, a4paper]{report}

@@

    \chapter{The Argennon Virtual Machine}\label{ch:AVM}

@@

    The Argennon Virtual Machine (AVM) is an abstract computing machine for executing Argennon's smart contracts. The
    Argennon Virtual Machine knows nothing of the Argennon blockchain, only of cleanup sessions and
    execution sessions.

    Execution sessions are separate sessions of executing smart contact's code by the Argennon Virtual Machine.
    These sessions usually are correspondent to the concept of a transaction in a blockchain. (See
    Section~\ref{subsec:execution-sessions})

    A cleanup session is the time slot in which the Argennon Virtual Machine performs cleanup and re-optimization of
    its internal data structures in a "Stop the World" manner. Cleanup sessions usually are done as the final phase of
    the block confirmation process in a blockchain. (See Section ...)

    \section{Data Types}\label{sec:data-types}

@@

    The Argennon Virtual Machine operates on two kinds of types: primitive types and reference types. There are,
    correspondingly, two kinds of values that can be stored in memory locations, passed as arguments, returned by
    methods, and operated upon: primitive values and reference values.

    Values of primitive or reference types need not be tagged or otherwise be inspectable to determine their types
    at run time, or to be distinguished from values of other types. Instead, the instruction set of the Argennon
    Virtual Machine distinguishes its operand types using instructions intended to operate on values of specific
    types. For instance, \texttt{iadd64} assumes that its operands are two 64-bit signed integers or \texttt{rstore}
    assumes that its operand is a reference type.

    A reference value may be the special \texttt{null} reference, a reference to no location, which will be denoted here
    by \texttt{null}. The default value of a reference type is \texttt{null}.

    The Argennon virtual machine does not have a fixed word size. Any instruction of the Argennon Virtual Machine has
    its specific word size, and the addressable memory areas are byte addressable.

    \section{Arithmetics}\label{sec:arithmetics}

    All arithmetic operations in the Argennon Virtual machine are safe and there are strict overflow checks and
    blah blah blah

    \section{Architecture}\label{sec:arch}

    \subsection{The \texttt{pc} Register}\label{subsec:the-pc-register}

    The Argennon Virtual Machine always has a single thread of execution and exactly has one \texttt{pc} register.
    When the Argennon Virtual Machine is executing a method, if that method is not native, the \texttt{pc} register
    contains the address of the AVM instruction currently being executed. If the method currently being executed
    is native, the value of the Argennon Virtual Machine's \texttt{pc} register is undefined.

    \subsection{Call Stack Queue}\label{subsec:call-stack}

    Every AVM execution session has a queue of call stacks. A call stack contains all the information that is needed for
    restoring the state, and continuing the execution after the method invocation completes.
    This information is represented by a \texttt{CallInfo} struct:
    \begin{verbatim}
CallInfo {
    applicationID,
    methodID,
    pc,
    canModify,
    localFrame,
    operandStack
}
    \end{verbatim}
    The \texttt{applicationID} field is the unique identifier that the AVM assigns to every smart contract,
    \texttt{methodID} is the unique identifier of a method's bytecode and \texttt{canModify} is a flag
    indicating whether a method is allowed to execute state modifying instructions or not.

    Every method invocation has a corresponding \texttt{CallInfo} struct and when a method is invoked its
    \texttt{CallInfo} struct is pushed onto the call stack. Hence, the top of the call stack always contains
    the \texttt{CallInfo} struct of the current method. When a method invocation completes its call
    information is popped from the call stack.

    An AVM execution session will continue as long as there is a non-empty call stack in the call stack queue. When the
    execution of the current call stack finishes the execution of the next call stack in the call stack queue starts.
    It's possible that an AVM method invocation creates new call stacks. These newly created call stacks will be a part
    of the current execution session and are added at the end of the call stack queue. A newly created call stack always
    contains a single \texttt{CallInfo} struct. See Section~\ref{subsec:method-invocation} for more details.

    \subsection{Run-Time Data Areas}\label{subsec:data-areas}

    The Argennon Virtual Machine defines various run-time data areas that are used during an execution session
    (i.e.\ transaction), Some of these data areas are persistent and are stored on the blockchain, Other data areas
    are per execution session or per method invocation. These data areas are created when their context starts and
    destroyed when their context ends.

    The Argennon Virtual Machine has five run-time data areas:
    \begin{itemize}
        \item Method Area
        \item Constant Area
        \item Local Frame
        \item Operand Stack
        \item Heap
    \end{itemize}
    All data areas except operand stack, have their own address space. Operand stack is a
    last-in-first-out (LIFO) stack and is not addressable. Every AVM memory access instruction operates on its specific
    data areas.

    \subsubsection{Method Area}

    The method area contains the byte-code of any method which can be called by a non-native method invocation
    instruction. In the Argennon Virtual Machine every method has a unique identifier:
    \texttt{methodID}, and the method area is a map from method identifiers to their byte codes.

    Instructions that modify the code area can only be run in privileged mode which means they can only be executed
    by the \emph{root} smart contract. As a result, the access of smart contracts to the code area is essentially
    read-only.

@@

    Every smart contract has a single constant area which is stored in the AVM code area as a special type of
    non-executable method. The constant area of a smart contract contains several kinds of constants, ranging from user
    defined constants to method address tables. A method address table stores the list of method identifiers of a smart
    contract and their access type. The access type of a method can be either \texttt{external} or \texttt{internal}.
    Only external methods can be invoked by \texttt{invoke\_external} instruction.

    \note{Instructions that modify the constant area can only be run in privileged mode.}

@@

    Every time a local frame is created, a corresponding empty last-in-first-out (LIFO) stack is created too. The
    Argennon Virtual Machine is a stack machine and its instructions take operands from the operand stack, operate on
    them, and push the result back onto the operand stack. An operand stack is destroyed when its owner method
    completes, whether that completion is normal or abrupt.

@@

    The heap of the Argennon Virtual Machine is a persistent memory area which stores \emph{memory chunks}. A Memory
    chunk is a byte addressable piece of memory which has a separate address space starting from 0. Each chunk has a
    fixed size and different chunks need not be equally sized. Chunks are stored in the heap area and every chunk is
    assigned a unique identifier: \texttt{chunkID}. As a result, the address of every memory location inside
    the heap area can be considered as a pair: \texttt{(chunkID, offset)}.

    A smart contract can read any chunk stored in the heap by having its \texttt{chunkID}. However, it can only modify
    those chunks that was created by itself. In other words, every memory chunk in the AVM heap has an owner. Only
    the owner can modify a chunk while any one can read that chunk. In addition, for modifying a chunk the \texttt{canModify}
    flag of the current method's \texttt{callInfo} struct must be true.

    \note{The reason behind this type of access control design is the fact that smart contract code is usually
    immutable. That means if a smart contract does not implement a getter mechanism for some parts of its internal
    data, this functionality can never be added later. Although the internal data is publicly available, there will
    be no way for other smart contracts to use this data on-chain. This design eliminates the need for implementing
    \emph{trivial} getters.}

    The AVM Heap is able to save snapshots of its state and later restore them. This will enable the Argennon
    Virtual Machine to have state reversion capability. The snapshot management of the heap area is done by
    the following functions:
    \begin{itemize}
        \item \texttt{Save()} saves a snapshot of the current state of the heap and pushes it onto the
        snapshot stack.
        \item \texttt{Restore()} pops the snapshot stored on the top of the snapshot stack and restores it. The
        current state of the heap will be lost.
        \item \texttt{Discard()} pops the snapshot stored on the top of the snapshot stack and discards it. The current
        state of the heap will be preserved.
    \end{itemize}

    \note{These functions are internal functions of the AVM. They are not instructions, and can not be called by smart
    contracts.}

    \section{Instruction Set Overview}\label{sec:instruction-set-overview}

    An Argennon Virtual Machine instruction consists of a \textbf{one-byte} opcode specifying the operation to be
    performed, followed by zero or more operands supplying arguments or data that are used by the operation.
    The number and size of the operands are determined solely by the opcode.

@@

    The Argennon Virtual Machine has four types of method invocation instruction:
    \begin{itemize}
        \item \texttt{invoke\_internal} invokes a method without changing the context of method execution. In other
        words, \texttt{applicationID} and \texttt{canModify} fields of the invoked method in the
        call stack will be the same as the invoker.
        A smart contract can invoke any method by \texttt{invoke\_internal} instruction, even if that method is an
        internal method of another smart contract. Since the invoked method will always be executed in the context of
        the invoking smart contract, a smart contract will not be able to modify another smart contract state by
        this instruction. This instruction facilitates the usage of libraries and code reuse.
        \item \texttt{invoke\_external} invokes a method and changes the context of method execution to another
        smart contract. Only external methods of another smart contract can be called by this instruction. For changing
        the context of method execution this instruction sets the \texttt{applicationID} field of the invoked method
        to the \texttt{applicationID} of the called smart contract. Then it searches the call stack and if
        the \texttt{applicationID} of the called smart contract is found in the call stack it sets \texttt{canModify}
        to \texttt{false}, otherwise \texttt{canModify} is set to \texttt{true}. As a result, if the
        called smart contract is already called, the invoked method will not be able to modify the heap area.
        \item \texttt{invoke\_native} invokes a method that is not hosted by the Argennon Virtual Machine. By this
        instruction, high performance native methods of the hosting machine could become available to AVM smart
        contracts. This instruction will not modify the AVM state.
        \item \texttt{spawn} spawns a new method invocation. Spawning a method invocation is the invocation of a method
        with changing the context of method execution and creating a new call stack. The created call stack is added at
        the end of the call stack queue and the \texttt{canModify} field of the \texttt{CallInfo} struct will always
        be set to \texttt{true}. This instruction does not modify the \texttt{pc} register and the execution of the
        caller will continue after this instruction. As a result, \texttt{spawn} can not return
        any value to the caller and the return value of the called method (if any) will be added
        to the transaction output. If the invoked method completes abruptly it will cause all the call stack queue to
        complete abruptly.
    \end{itemize}

    \note{The AVM does not natively support polymorphism and virtual methods. A compiler could easily
    generates appropriate code for implementing these features.}

    Each time a method is invoked the \texttt{Save()} function of the AVM heap is called, then a new local frame
    and operand stack is created. The Argennon Virtual Machine uses local frames to pass parameters on
    method invocation. On method invocation, any parameters are passed in consecutive local variables stored in the
    method's local frame starting from address 0. The invoker of a method writes the parameters in the local frame
    of the invoked method using \texttt{arg} instructions.

    \note{For AVM smart contracts \emph{reentrancy} can only happen in read-only mode and is totally safe.
    At the same time call-back patterns can be easily implemented by using \texttt{spawn} instruction.}

    \subsection{Exceptions}\label{subsec:exceptions}

    An exception is thrown programmatically using the \texttt{athrow} instruction. Exceptions can also be thrown by
    various Argennon Virtual Machine instructions if they detect an abnormal condition. Some exceptions are not
    catchable and will always abort the method invocation.

@@

    \subsection{Method Invocation Completion}\label{subsec:method-invocation-completion}

    A method invocation completes normally if that invocation does not cause an exception to be thrown, either
    directly from the AVM or as a result of executing an explicit throw statement. If the invocation of the current
    method completes normally and the invocation was made by an \texttt{invoke} instruction, then a value may be
    returned to the invoking method. This occurs when the invoked method executes one of the return instructions, the
    choice of which must be appropriate for the type of the value being returned (if any). Execution then continues
    normally in the invoking method's local frame with the returned
    value (if any) pushed onto the operand stack. If the method was invoked by a \texttt{spawn} instruction,
    the execution of the next call stack (if any) begins and the returned value
    is appended at the end of the transaction output. When a method invocation completes normally
    the \texttt{Heap.Discard()} function is called as a part of the return instruction.

    A method invocation completes abruptly if an exception is thrown and is not caught by the current method. A
    method invocation that completes abruptly never returns a value.

    When a method completes, whether normally or abruptly, the call stack is used to restore the state of the invoker,
    including its local frame and operand stack, with the \texttt{pc} register appropriately restored and incremented
    to skip past the method invocation instruction. If the current call stack is empty the next call stack in the call
    stack queue will be used for continuing the execution, and when there are no call stacks left in the queue the
    current execution session ends.

    A thrown exception causes methods in the call stack to complete \emph{abruptly} one by one, as long as the
    \texttt{pc} register is not pointing to a \texttt{catch} instruction. The \texttt{catch} instruction acts like a
    branch instruction that branches only if an exception is caught. When an exception is thrown
    Algorithm~\ref{alg:Abrupt} is used to restore the state of the AVM heap.

    \begin{algorithm}
        \DontPrintSemicolon
        \SetKwData{Heap}{Heap}\SetKwData{Stack}{CallStack}
        \SetKwFunction{Discard}{Discard}\SetKwFunction{Pop}{pop}\SetKwFunction{Restore}{Restore}
        \SetKwProg{Fn}{Function}{}{}

        \While{\Stack is not empty}
        {
            \Stack.\Pop{}\;
            \eIf{\Stack is not empty \textbf{and} $\Stack[head].pc = catch$}
            {
                \Heap.\Restore{}\;
            }{
                \Heap.\Discard{}\;
            }
        }
        \Heap.\Restore{}\;
        \caption{Abrupt Method Completion}\label{alg:Abrupt}
    \end{algorithm}

    Closer investigation of Algorithm~\ref{alg:Abrupt} shows that when an invocation which was made by
    a \texttt{spawn} instruction completes abruptly, the heap snapshot that was saved before the start of the
    execution of the first call stack of the execution session is restored. This means
    that all changes made to the AVM heap during the execution session will be restored, but the changes made by
    the root smart contract including the transferring of the transaction fee will not be restored.
    See Section~\ref{subsec:execution-sessions} for more details.

    \subsection{Heap Allocation and De-Allocation}\label{subsec:heap-allocation-instructions}

    \section{Authorizing Operations}\label{sec:authorizing-operations}

    In blockchain applications, we usually need to authorize certain operations. For example, for sending an asset
    from a user to another user, first we need to make sure that the sender has authorized this operation. The
    Argennon virtual machine has no built in mechanism for authorizing operations, but it provides a rich set of
    cryptographic instructions for validating signatures and cryptographic entities. By using these instructions and
    passing cryptographic signatures as parameters to methods, a programmer can implement the required logic
    for authorizing any operation.

@@

    In addition to signatures, a method can verify its invoker by using \texttt{get\_parent} instruction. This
    instruction gets the \texttt{applicationID} of the smart contract that is one level deeper than the current
    smart contract in the call stack or if there is none, the smart contract that created the current call stack.
    In other words, it returns the \texttt{applicationID} of the smart contract that has invoked or spawned
    the current smart contract.

    \section{Execution Sessions}\label{subsec:execution-sessions}

    The execution of smart contracts code in the AVM is done in execution sessions. Every execution session in the
    Argennon Virtual Machine constitutes of a single method invocation from the \emph{root} smart contract and a list
    of resource allocation requests. These allocation requests always include a request for allocating some amount
    of execution time and a request for accessing some memory locations for reading or writing. If after the start of
    execution an execution session tries to violate its allocated resources, an uncatchable exception will
    be thrown by the AVM\@.

    \note{The AVM does not have a \texttt{calldata} memory area. The arguments of the root smart contract's method will
    be copied from the transaction to the local frame of the invoked method, exactly like a normal method invocation.}

    The root smart contract with the \texttt{applicationID} of zero, is a special smart contract in the Argennon
    Virtual Machine. Its code is immutable and a part of the AVM implementation and always is run in privileged mode.
    Every execution session starts with a call to a method of the root smart contract. This method always transfers
    some amount of fee from an account ot the \texttt{feeSink} account and then it performs the requested operation.

    If the requested operation is a method invocation the root smart contract will call the \texttt{Heap.Save()}
    function once and then it will invoke the requested method using \texttt{spawn} instruction. This means that the
    invoked method will have a separate call stack and if it throws an exception it will not cause a rollback in
    changes the is made by the root smart contract to the state.

    \note{Argennon transactions do not have a sender, and the payer of a transaction fee could be any
    account who has provided the required digital signature.}

    \section{Cleanup Sessions}\label{sec:cleanup-sessions}

@@

    \chapter{The Argennon Blockchain}\label{ch:argennon-blockchain}
